### 2017-12-20 chapters/introduction/mechanics.tex
Add latest version of summary section

In this section we will cover the necessary theory of continuum
mechanics in order to model the mechanics of the heart. The theory of
continuum mechanics is extensive, and we will not be able to cover
everything. For a complete review of continuum mechanics the reader is
therefore referred the textbook of Gerard Holzapfel
\cite{holzapfel2000nonlinear} from which most of the theory in this
section is taken. For the even more mathematically oriented reader we
refer to \cite{marsden1994mathematical}.

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The mapping $\varphi$ will be
referred to as the \emph{motion}. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of the motion with
respect to the material coordinates:
\begin{align}
  \F = \nabla_{\Xvec} \varphi= \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}

@@

\subsubsection{Transformations between reference and current
  configuration}
By definition, the reference configuration $\Omega$, and current
configuration $\omega$, are related via the motion $\varphi$ in the
sense that a point $\mathfrak{p} \in \mathfrak{B}$ with reference
coordinates $\Xvec$ and current coordinates $\xvec$ satisfies $\xvec =
\varphi(\Xvec)$. Likewise a vector in the reference configuration is
related to a vector in the current configuration  via the
deformation gradient $\F$; if $\mathrm{d}\Xvec$ is a vector in the
reference configuration it will transform to the vector
$\mathrm{d}\xvec$ in the current configuration, and $\mathrm{d}\xvec =
\F \mathrm{d}\Xvec$. From this relation we also derive that the
transformation of an infinitesimal volume element in the reference
configuration, $\mathrm{d}V$ is related to an infinitesimal volume
element in the current configuration, $\mathrm{d}v$  via the determinant of the
deformation gradient,
\begin{align}
  \mathrm{d}v =\det(\F) \mathrm{d}V.
  \label{eq:volume_element}
\end{align}
Another important transformation is the transformation of normal
vectors. By noting that we can write \eqref{eq:volume_element} using
surface elements
\begin{align*}
  \mathrm{d}s \mathbf{n} \mathrm{d}\xvec  &= \mathrm{d}v = \det(\F) \mathrm{d}V = \det(\F) \mathrm{d}S  \Nvec \mathrm{d}\Xvec\\
  &\implies \left( \mathrm{d}s \mathbf{n} \F  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0\\
  &\implies \left( \mathrm{d}s \F^T \mathbf{n}  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0,\\
\end{align*}
we get \emph{Nanson's formula}
\begin{align}
  \mathrm{d}s \mathbf{n}  =  \det(\F) \F^{-T} \mathrm{d}S \Nvec,
\end{align}
which relates the normal vector in the current configuration to the
normal vector in the reference configuration.

\subsubsection{Conservation of linear momentum}
Newton's seconds law states that the change in linear momentum equals
the net impulse acting on it. For a continuum material with constant
mass density $\rho$ this implies that
\begin{align}
  \int_{\omega} \rho \dot{\mathbf{v}} \mathrm{d}v = \mathbf{f},
  && \mathbf{f} = \int_{\partial \omega} \mathbf{t} \mathrm{d}s
     + \int_{\omega} \mathbf{b} \mathrm{d}v,
     \label{eq:cons_lin_mom}
\end{align}
where $\mathbf{v}$ is the spatial velocity field, $\mathbf{t}$ is
the traction acting on the boundary, and $\mathbf{b}$ is the body
force. From \emph{Cauchy's stress theorem} we know that there exists a
second order tensor $\sigma$, known as the Cauchy stress tensor that is
related to the traction vector by $\mathbf{t} = \sigma \mathbf{n}$,
where $\mathbf{n}$ is the unit normal in the current configuration.
Using the divergence theorem we get
\begin{align*}
  \int_{\partial \omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom} we arrive at
Cauchy's momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) can be considered negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor.

@@

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is, the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T = \mathbf{a}_0\otimes\mathbf{a}_0 \},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
The invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.

>>>>

In this section we will cover the necessary theory of continuum
mechanics in order to model the mechanics of the heart. The theory of
continuum mechanics is extensive, and we will not be able to cover
everything. For a complete review of continuum mechanics the reader is
therefore referred the textbook of Gerhard Holzapfel
\cite{holzapfel2000nonlinear} from which most of the theory in this
section is taken. For the even more mathematically oriented reader we
refer to \cite{marsden1994mathematical}.

\subsection{Kinematics}
We represent the heart as a continuum body $\mathfrak{B}$ embedded in
$\mathbb{R}^3$. A configuration of $\mathfrak{B}$ is a mapping $\chi:
\mathfrak{B} \rightarrow \mathbb{R}^3$. 
We denote the \emph{reference configuration} of the heart by $\Omega
\equiv \chi_0(\mathfrak{B})$, and the \emph{current configuration} by $\omega
\equiv \chi(\mathfrak{B})$. The mapping $\varphi :  \Omega
\rightarrow \omega$, given by the composition $\varphi = \chi
\circ \chi_0^{-1}$, is a smooth, orientation preserving (positive
determinant) and invertible map. We denote the coordinates in the
reference configuration by $\Xvec \in \Omega$, and the coordinates in the current
configuration by $\xvec \in \omega$. The coordinates $\Xvec$ and $\xvec$ are
commonly referred to as material and spatial points respectively, and
are related through the mapping $\varphi$, by $\xvec = \varphi(\Xvec)$.
For time-dependent problems it is common to make  the time-dependence
explicitly by writing $\xvec = \varphi(\Xvec, t)$. In the following
we will only focus on the mapping between two configurations and
therefore no time-dependence is needed. The \emph{deformation gradient} is a
rank-2 tensor, defined as the partial derivative of $\varphi$  with
respect to the material coordinates:
\begin{align}
  \F = \nabla_{\Xvec} \varphi= \Grad \xvec.
  \label{eq:deformation_gradient}
\end{align}
Here we also introduce the notation $\Grad$, which means derivative
with respect to reference coordinates.
The deformation gradient maps vectors in the reference configuration to
vectors in the current configuration, and belongs to the space of
linear transformations from $\mathbb{R}^3$ to $\mathbb{R}^3$ with
strictly positive determinant, which we denote by
$\mathrm{Lin}^+$. Another important quantity is the
\emph{displacement} field  
\begin{align}
  \uvec = \xvec-\Xvec, 
  \label{eq:displacement}
\end{align}
which relates positions in the reference configuration to positions
in the current configuration. From \eqref{eq:deformation_gradient} we
see that
\begin{align}
  \F = \Grad \xvec = \Grad \uvec + \Grad \Xvec = \Grad \uvec + \I.
\end{align}

@@

\subsubsection{Transformations between reference and current
  configuration}
By definition, the reference configuration $\Omega$, and current
configuration $\omega$, are related via the deformation map $\varphi$ in the
sense that a point $\mathfrak{p} \in \mathfrak{B}$ with reference
coordinates $\Xvec$ and current coordinates $\xvec$ satisfies $\xvec =
\varphi(\Xvec)$. Likewise a vector in the reference configuration is
related to a vector in the current configuration  via the
deformation gradient $\F$; if $\mathrm{d}\Xvec$ is a vector in the
reference configuration it will transform to the vector
$\mathrm{d}\xvec$ in the current configuration, and $\mathrm{d}\xvec =
\F \mathrm{d}\Xvec$. From this relation we also derive that the
transformation of an infinitesimal volume element in the reference
configuration, $\mathrm{d}V$ is related to an infinitesimal volume
element in the current configuration, $\mathrm{d}v$  via the determinant of the
deformation gradient,
\begin{align}
  \mathrm{d}v =\det(\F) \mathrm{d}V.
  \label{eq:volume_element}
\end{align}
Another important transformation is the transformation of normal
vectors. By noting that we can write \eqref{eq:volume_element} using
surface elements
\begin{align*}
  \mathrm{d}s \mathbf{n} \mathrm{d}\xvec  &= \mathrm{d}v = \det(\F) \mathrm{d}V = \det(\F) \mathrm{d}S  \Nvec \mathrm{d}\Xvec\\
  &\implies \left( \mathrm{d}s \mathbf{n} \F  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0\\
  &\implies \left( \mathrm{d}s \F^T \mathbf{n}  - \mathrm{d}S \det(\F) \Nvec \right) \mathrm{d}\Xvec = 0,\\
\end{align*}
we get \emph{Nanson's formula}
\begin{align}
  \mathrm{d}s \mathbf{n}  =  \det(\F) \F^{-T} \mathrm{d}S \Nvec,
\end{align}
which relates the normal vector in the current configuration to the
normal vector in the reference configuration.

\subsubsection{Conservation of linear momentum}
Newton's seconds law states that the change in linear momentum equals
the net impulse acting on it. For a continuum material with constant
mass density $\rho$ this implies that
\begin{align}
  \int_{\omega} \rho \dot{\mathbf{v}} \mathrm{d}v = \mathbf{f},
  && \mathbf{f} = \int_{\partial \omega} \mathbf{t} \mathrm{d}s
     + \int_{\omega} \mathbf{b} \mathrm{d}v,
     \label{eq:cons_lin_mom}
\end{align}
where $\mathbf{v}$ is the spatial velocity field, $\mathbf{t}$ is
the traction acting on the boundary, and $\mathbf{b}$ is the body
force. From \emph{Cauchy's stress theorem} we know that there exists a
second order tensor $\sigma$, known as the Cauchy stress tensor that is
related to the traction vector by $\mathbf{t} = \sigma \mathbf{n}$,
where $\mathbf{n}$ is the unit normal in the current configuration.
Using the divergence theorem we get
\begin{align*}
  \int_{\partial \omega} \mathbf{t} \mathrm{d}s
  = \int_{\partial \omega} \sigma \mathbf{n} \mathrm{d}s
  = \int_{\omega} \nabla \cdot \sigma \mathrm{d}v,
\end{align*}
and by collecting the terms from \eqref{eq:cons_lin_mom} we arrive at
Cauchy's momentum equation
\begin{align}
  \nabla \cdot \sigma + \mathbf{b} =  \rho \dot{\mathbf{v}}.
  \label{eq:chauch_momentum_eq}
\end{align}
The contribution from the body force ($\mathbf{b}$)  and inertial term
($\rho \dot{\mathbf{v}}$) can be considered negligible compared to the stresses
\cite{hunter1996kd,tallarida1970left, moskowitz1981effects}, which is
why the force balance equations is typically only stated as
\begin{align}
  \nabla \cdot \sigma = \mathbf{0}.
  \label{eq:momentum_simple_current}
\end{align}
Note that we have formulated the balance law in the current
configuration. An equivalent statement can be formulated in terms of
the reference configuration
\begin{align}
  \nabla \cdot \FPK = \mathbf{0}, 
  \label{eq:momentum_simple_reference}
\end{align}
where $\FPK$ is the first Piola-Kirchhoff stress tensor. Note that the
operator $\nabla \cdot$ acting on the Cauchy stress tensor represents
differentiation with respect to coordinates in the current
configuration, while when acting on the first Piola-Kirchhoff stress
tensor represent differentiation with respect to coordinates in the
referece configuration.

@@

\subsubsection{General requirements for the strain-energy density
  function}
\label{sec:strain_energy_req}
Some general requirements must hold for the strain-energy function.
First of all, we require that the reference state is stress free and
that the stored energy increases monotonically with the deformation. 
Formally this can be stated simply as 
\begin{align*}
  \Psi(\I) = 0 \; && \text{and} &&\; \Psi(\F) \geq 0.
\end{align*}
Moreover, expanding or compressing a body to zero volume would
require an infinite amount of energy, i.e
\begin{align*}
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& 0 \\
  \Psi(\F) \rightarrow \infty \; && \text{as} &&\; \det \F \rightarrow& \infty
\end{align*}
We say that the strain energy should be objective, meaning that the
stored energy in the material should be invariant with respect to
change of observer. Formally we must have: \emph{given any positive symmetric
rank-2 tensor $\C \in \mathrm{Sym}$:}
\begin{align}
  \Psi(\mathbf{C}) = \Psi(\mathbf{Q}\mathbf{C}\mathbf{Q}^T), \; \forall \mathbf{Q} \in \mathcal{G} \subseteq \mathrm{Orth}.
\end{align}
Here $\mathrm{Orth}$ is the group of all positive orthogonal matrices.
If $\mathcal{G} = \mathrm{Orth}$ we say that the material is
isotropic, and otherwise we say that the material is anisotropic.
This brings us to another important issue, which is related to the
choice of coordinate-system. Having to deal with different
coordinate-systems, and mapping quantities from one coordinate-system
to another can results in complicated computations. Therefore it would be beneficial if we
could work with quantities which do not depend on the choice of
coordinate-system. Such quantities are called invariants. 
If the material is isotropic, the representation theorem for
invariants states that $\Psi$ can be expressed in terms of the
principle invariants of $\mathbf{C}$, that is $\Psi = \Psi(I_1, I_2,
I_3)$. The principle invariants $I_i, i=1,2,3$ are the coefficients in
the characteristic polynomial of $\mathbf{C}$, and is given by 
\begin{align}
  I_1 = \tr \C,  && I_2 = \frac{1}{2}\left[ I_1^2 - \tr(\C^2)\right] && \text{and} && I_3 = \det \C.
\end{align}
In the case when the material constitutes a transversely isotropic
behavior, that is, the material has a preferred direction $\mathbf{a}_0$,
which in the case of the myocardium could be the direction of fiber
muscle fibers, we have
\begin{align*}
  \mathcal{G} = \{ \mathbf{Q} \in \mathrm{Orth}: \mathbf{Q}(\mathbf{a}_0\otimes\mathbf{a}_0)\mathbf{Q}^T = \mathbf{a}_0\otimes\mathbf{a}_0 \},
\end{align*}
with $\otimes$ being the outer product. In this case the strain-energy
density function can still be expressed through invariants. However,
we need to include the so called quasi-invariants, which are defined
as stretches in the local microstructural coordinate-system. The
transversely isotropic invariants are given by
\begin{align*}
  I_{4\mathbf{a}_0 } = \mathbf{a}_0 \cdot (\C \mathbf{a}_0) && \text{and} && I_5 = \mathbf{a}_0 \cdot (\C^2 \mathbf{a}_0).
\end{align*}
Some of the invariants do have a physical interpretation. For instance, $I_3$
is related to the volume ratio of material during deformation, while
$I_{4\mathbf{a}_0 } $ is related to the stretch along the direction
$\mathbf{a}_0 $. Indeed the \emph{stretch} ratio in the direction
$\mathbf{a}_0$ is given by $\lambda_{\mathbf{a}_0} = | \F \mathbf{a}_0
|$ and we see that $I_{4\mathbf{a}_0 }  =  \mathbf{a}_0 \cdot (\C
\mathbf{a}_0) = \F \mathbf{a}_0 \cdot (\F \mathbf{a}_0) =
\lambda_{\mathbf{a}_0}^2$. For more details about invariants see e.g
\cite{holzapfel2009constitutive,liu1982representations}.
